---
title: Endangered Plants
author: Joseph Pope
date: '2020-08-18'
slug: endangered-plants
categories:
  - R
tags:
  - tidyverse
comments: yes
description: 'A TidyTuesday project exploring endangered plant life.'
images:
  - src: 'img/2020/tidytuesday/clearcut.jpg'
    stretch: stretchH
removeBlur: no
---

Today we are analyzing extinct and endangered plants for TidyTuesday. 

Every week I find myself starting my research path at the truly excellent site [Data-to-Viz by Yan Holtz and Conor Healy](https://www.data-to-viz.com/) for inspiration on what I want to try and learn with R. The site is interactive and helps you figure out the best visualization given numerical or categorical data. It's well worth a bookmark in your favorite browser. 

This week's dataset has three tables, many of which have several columns/features and ways in which to cut up the data. I decided to try something I've never done before, and build a [Circular Barplot](https://www.data-to-viz.com/graph/circularbarplot.html). 

The code is long and complicated. I do not recommend this for beginner's. However, the effect is pretty striking and it's a nice weapon to have in the belt. I was tempted to build a stacked circular bar chart, which is REALLY impressive, but for my first foray into polar coordinates, I decided to just do a grouped bar chart, which honestly was pretty tough on its own. 


Setup 

```
# Contents of Setup.R. 
# Using Source method to suppress warnings and messages from Rmd output

# library(tidyverse)
# library(extrafont)
# library(ggthemes)
# library(tidyverse)
# library(knitr)
# library(kableExtra)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
source('../../setup.R')

# Contents of Setup.R. 
# Using Source method to suppress warnings and messages from Rmd output

# library(tidyverse)
# library(extrafont)
# library(ggthemes)
# library(tidyverse)
# library(knitr)
# library(kableExtra)

knitr::opts_chunk$set(
	fig.height = 10,
	fig.width = 10,
	message = FALSE,
	warning = FALSE,
	dpi = 600
)
```


# Load Data
```{r}
# tuesdata <- tidytuesdayR::tt_load('2020-08-18')

# threats <- tuesdata$threats

threats <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-08-18/threats.csv')
```

# Explore Data 
```{r}
skimr::skim(threats)
```

# Explore Data MORE
```{r}
table(threats$threat_type, threats$threatened)

```

# Data Processing
```{r}

data <- threats %>% 
  filter(threatened == 1) %>% 
  group_by(continent, country) %>% 
  count(name = "threatened_plants", sort = TRUE) %>% 
  ungroup() %>% 
  mutate(continent = factor(continent),
         country = ifelse(str_detect(country, "Saint Helena"), "Saint Helena", country),
         threatened_plants = scales::rescale(threatened_plants, to = c(0, 100))) %>% 
  filter(threatened_plants != 0)
```

# Plotting

```{r}
# Set a number of 'empty bar' to add at the end of each group
empty_bar=3
to_add = data.frame( matrix(NA, empty_bar*nlevels(data$continent), ncol(data)) )

colnames(to_add) = colnames(data)
to_add$continent = rep(levels(data$continent), each=empty_bar)

data=rbind(data, to_add)
data=data %>% arrange(continent)
data$id=seq(1, nrow(data))


# Get the name and the y position of each label
label_data=data
number_of_bar=nrow(label_data)
angle= 90 - 360 * (label_data$id-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
label_data$hjust<-ifelse( angle < -90, 1, 0)
label_data$angle<-ifelse(angle < -90, angle+180, angle)


# prepare a data frame for base lines
base_data=data %>% 
  group_by(continent) %>% 
  summarize(start=min(id), end=max(id) - empty_bar) %>% 
  rowwise() %>% 
  mutate(title=mean(c(start, end)))

# prepare a data frame for grid (scales)
grid_data = base_data
grid_data$end = grid_data$end[ c( nrow(grid_data), 1:nrow(grid_data)-1)] + 1
grid_data$start = grid_data$start - 1
grid_data=grid_data[-1,]

```


# Build Plot

```{r}

p = ggplot(data, aes(x=as.factor(id), y=threatened_plants, fill=continent)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  
  geom_bar(aes(x=as.factor(id), y=threatened_plants, fill=continent), stat="identity", alpha=0.5) +
  
  # Add a val=100/75/50/25 lines. I do it at the beginning to make sur barplots are OVER it.
  geom_segment(data=grid_data, aes(x = end, y = 80, xend = start, yend = 80), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 60, xend = start, yend = 60), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 40, xend = start, yend = 40), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
  geom_segment(data=grid_data, aes(x = end, y = 20, xend = start, yend = 20), colour = "grey", alpha=1, size=0.3 , inherit.aes = FALSE ) +
  
  # Add text showing the value of each 100/75/50/25 lines
  annotate("text", x = rep(max(data$id),4), y = c(20, 40, 60, 80), label = c("20", "40", "60", "80") , color="grey", size=3 , angle=0, fontface="bold", hjust=1) +
  
  geom_bar(aes(x=as.factor(id), y=threatened_plants, fill=continent), stat="identity", alpha=0.5) +
  ylim(-100,120) +
  labs(
    title = "Number of Threatened Plants by Geo",
    subtitle = "Scaled from 0-100",
    caption = "Joseph Pope | @joepope44"
  ) + 
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 24, family  = "Garamond"),
    plot.subtitle = element_text(hjust = 0.5, family  = "Garamond"),
    plot.caption = element_text(hjust = 0.5, family = "Garamond")
    # plot.margin = unit(rep(-1,4), "cm") <- this hides the plot title if not commented out
  ) +
  coord_polar() + 
  geom_text(data=label_data, aes(x=id, y=threatened_plants+10, label=country, hjust=hjust), color="black", fontface="bold",alpha=0.7, size=3.5, angle= label_data$angle, inherit.aes = FALSE ) +
  
  # Add base line information
  geom_segment(data=base_data, aes(x = start, y = -5, xend = end, yend = -5), colour = "black", alpha=0.8, size=0.6 , inherit.aes = FALSE )  +
  geom_text(data=base_data, aes(x = title, y = -18, label=continent),
            hjust=c(1,1,.5,0,0,0), colour = "black", alpha=0.8, size=3, 
            fontface="bold", inherit.aes = FALSE) 
 
p
```

